---
title: "Tint Is Not Tufte: Brief Intro"
author: "Dirk Eddelbuettel and Jonathan Gilligan"
date: "`r Sys.Date()`"
output: tint::tintHtml
bibliography: tint.bib
link-citations: yes

vignette: >
  %\VignetteIndexEntry{Short tint HTML Example}
  %\VignetteKeywords{tint,vignette}
  %\VignettePackage{tint}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
library(tint)
library(ggplot2)
mtcars$am <- factor(mtcars$am, labels=c("manual", "automatic"))
## download files if connected
## when not connected, two margin figure will not be shown, the rest
## is still processed as usual
connected <- tint:::.isConnected()
if (connected) {
    tmpdir <- normalizePath(tempdir(), winslash = "/")
    latofile <- file.path(tmpdir, "tintPdfLatoPage1.png")
    garamondfile <- file.path(tmpdir, "tintPdfGaramondPage1.png")
    download.file("https://eddelbuettel.github.io/tint/tintPdfLatoPage1.png",
                  latofile, quiet = TRUE)
    download.file("https://eddelbuettel.github.io/tint/tintPdfGaramondPage1.png",
                  garamondfile, quiet = TRUE)

}
```

# An Introduction to `tint`

[tint](https://github.com/eddelbuettel/tint) [@cran:tint] updates the look and feel of
'Tufte' documents for R. It combines the (html and pdf parts of the) excellent
[tufte](https://github.com/rstudio/tufte) package [@cran:tufte] with the [Roboto
Condensed](https://fonts.google.com/specimen/Roboto+Condensed) font use and color scheme
proposed by [envisioned css](https://github.com/nogginfuel/envisioned-css) plus minor
style changes such as removal of _italics_---but remains otherwise true to the
[tufte](https://github.com/rstudio/tufte) package for R.  Later additions are a book style
and well as generalisation of the font handling allowing for
[Lato](http://www.latofonts.com/lato-free-fonts/) and
[Garamond](https://fonts.adobe.com/fonts/adobe-garamond) fonts along with extended color
selection.

```{r fig-nocap-margin-first, fig.margin=TRUE, fig.width=4, fig.height=4, cache=TRUE, echo=FALSE}
ggplot(mtcars, aes(wt, mpg)) + geom_point(size=3, aes(colour=factor(cyl))) +
    theme(legend.position="none")
```

The package name follows an old tradition and is recursive: _tint is not tufte_.

Full documentation is available in the [longer PDF
vignette](https://eddelbuettel.github.io/tint/tintPDF.pdf), its
[Lato](https://eddelbuettel.github.io/tint/tintPDFLato.pdf) and
[Garamond](https://eddelbuettel.github.io/tint/tintPDFGaramond.pdf) variants (see below)
as well as the [longer HTML vignette](https://eddelbuettel.github.io/tint/tintHTML.html).
As these render to about two megabytes each, we no longer include them by default in the
package as it swells the size of the installed package unnecessarily.

# Margin Examples

The margin can be used for arbirtrary 'figure' environments by using the `knitr` option
`marginfigure` along with standard \LaTeX mathematical markup.

```{marginfigure}
We know from _the first fundamental theorem of calculus_ that for $x$ in $[a, b]$:
$$\frac{d}{dx}\left( \int_{a}^{x} f(u)\,du\right)=f(x).$$
```

We can also add a 'note' using the standard `pandoc` notation of text in square brackets
following an hat symbol: `^[text here]`.^[This is a sidenote that was entered using a
footnote.]

Notes can also be added in-line using an R command and the `margin_note()` function. 
`r margin_note("Some text.")` Note that this note is unnumbered.

# Margin Figures

One feature of the [tint](https://github.com/eddelbuettel/tint) /
[tufte](https://github.com/rstudio/tufte) packages are margin figure such as the first one
on the right. It was created by setting the `knitr` option `fig.margin=TRUE`; the plotting
code itself is standard.

```r
ggplot(mtcars, aes(wt, mpg)) +
    geom_point(size=3, aes(colour=factor(cyl))) +
    theme(legend.position="none")
```


```{r fig-nocap-margin-second, fig.margin=TRUE, fig.width=4, fig.height=4, cache=TRUE, echo=FALSE}
ggplot(mtcars, aes(wt, mpg)) + geom_point(size=3, aes(colour=factor(cyl))) +
    theme_tint() + theme(legend.position="none")
```

Charts can also take advantage of the `theme_tint()` we added; it owes a lot to a similar
function in the [ggtufte](https://github.com/jrnold/ggtufte) package
[@github:ggtufte] (but which conflicts in its font settings with our, so we
simplified). Its effect can be seen in the second figure. Note that the legend-suppression
has to come after `theme_tint()` as changes are additive.



# Full Width Figures

Figures can span across the entire page; this is enabled by using the chunk option
`fig.fullwidth = TRUE`. Using the default them but conditioning by number of cylinders:

```{r fig-fullwidth, fig.width = 10, fig.height = 3, fig.fullwidth = TRUE, fig.cap = "A full width figure.", echo=FALSE, cache=TRUE}
ggplot(mtcars, aes(wt, mpg)) + geom_point(size=3, aes(colour=am)) + facet_wrap(~ factor(cyl)) +
    theme(legend.position="bottom")
```

# Main Column Figures

Besides margin and full width figures, one can of course also include figures constrained
to the main column. This is the default type of figures in the LaTeX/HTML output. A single
figure with cylinders in color and transmission not controlled for, and once again using
`theme_tint()`:

```{r fig-main, fig.cap = "A figure in the main column.", cache=TRUE, echo=FALSE}
ggplot(mtcars, aes(wt, mpg)) +
    geom_point(size=2, aes(colour=factor(cyl))) +
    theme_tint() + theme(legend.position="none")
```

# Font Extensions

Since version 0.1.1, the Lato and Garamond font families can be used.  See the package for
details. In contrast to the [default Roboto
variant](https://eddelbuettel.github.io/tint/tintPDF.pdf), the [Lato
variant](https://eddelbuettel.github.io/tint/tintPDFLato.pdf) uses

```{r fig-lato-screenshot, fig.margin=TRUE, cache=TRUE, echo=FALSE}
if (connected) knitr::include_graphics(latofile, dpi = NA)
```

```yaml
latexfonts: 
  - package: lato
    options: default
  - package: FiraMono
linkcolor: "0.3,0.3,0.6"
```


The [Garamond variant](https://eddelbuettel.github.io/tint/tintPDFGaramond.pdf) uses

```yaml
latexfonts: 
  - package: newtxmath
    options: 
      - cmintegrals
      - cmbraces
  - package: ebgaramond-maths
  - package: nimbusmononarrow
```

```{r fig-garamond-screenshot, fig.margin=TRUE, cache=TRUE, echo=FALSE}
if (connected) knitr::include_graphics(garamondfile, dpi = NA)
```

This requires the fonts to be installed on the system on which the document is
prepared. How to install additional fonts is beyond the scope of this note, see platform
and font-specific help.

# Citations, Code, Tables and More

## Citations

One can use Pandoc-style citations using the the Bibtext citation identifier inside of
square brackets: `[@someone:1984]`. In pdf mode, the standard `natbib` features are
also available such as `\citet{}`, `\citep{}` and more.

## Code

The package, just like any other `pandoc`-driven use of markdown, can also typset code
directly.  We showed this above with the `ggplot()` example (using R formatting) and the
font declarations (using YAML formatting).

## Tables

As [knitr](https://github.com/yihui/knitr)
[@cran:knitr] is driving the conversion, many of its options also apply.  See the
documentation for the [tufte](https://github.com/rstudio/tufte) and
[knitr](https://github.com/yihui/knitr) for details.

